### 2016-04-23 manuscript-body.tex
Add introduction and flesh out juice box pressure example

@@

@@

To be clear, it's not just that graphics might improve our efficiency in
debugging a statement like line 11. Rather, choosing to
use conceptual knowledge from physics to understand a piece of code transforms debugging a position-update statement
into a different activity: reasoning kinematically about the properties of motion
trails. For a comparative perspective, consider Hutchins's argument that for aircraft pilots, ringing an aircraft speedometer with physical markers transforms the cognition involved in landing a plane \cite{hutchins_how_1995}:

@@

In Hutchins's

the cognitive act of debugging in this instance. Applying an interpretive frame that
treats an assignment statement as a kinematic position update lets us
\emph{use conceptual knowledge from physics} to diagnose and fix
problems in code.   There,  For comparison, consider

@@

\subsection{Example 2: The Same Symbol Template Within The Same Language Can Still Mean Different Things}

@@

% \begin{supertabular}[]{@{}llll@{}}
% \toprule
% \textbf{Statement } & Programming Syntax & Language & How we think about
% it\tabularnewline
% \midrule
% \endhead
% \textbf{1} & i = i + 1 & C & \textbf{Increments} i by 1\tabularnewline
% \textbf{2} & w = w + ``ly'' & JavaScript & \textbf{Appends} ``ly'' to
% the string w\tabularnewline
% \textbf{3} & p = p + geom\_point() & R (ggplot2) & \textbf{Composes} a
% layer of points onto a plot p\tabularnewline
% \bottomrule
% \end{supertabular}

The catch here is that the plus operator takes on different roles in
different languages because of how those languages define its use.
Statement 1 increments a number in C; Statement 2 appends the letters
``ly'' to a string; Statement 3 composes a layer of points onto a statistical
graphics plot. These different kinds of operations become even more
apparent and consequential when, for example, such statements are
repeated in the same language within the same file. In the R/ggplot2 code below, the author uses multiple
reassignment statements to compose a statistical graphics plot.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Programming can be helped by applying conceptual models to code,
  particularly when relevant domain-knowledge structures can
  advantageously transform a problem (example 1)
\item
  But, conceptual models don't work all the time for all statements.
  Because languages are designed differently, the same syntax can
  actually correspond to very different operations in code (example 2).
  And, that's true both within and across languages.
\item
  Consequently, it makes less sense to treat conceptual models as right
  or wrong, and more sense to treat them as differentially advantageous
  for thinking about what a piece of code does. (Thinking a ``+''
  implies numerical addition isn't \emph{globally wrong} in JavaScript,
  but it won't explain why 1 + ``1'' yields ``11'' as a result.)
\item
  It seems plausible that successful programmers, when reasoning about
  or writing code, are able to dynamically access or deploy conceptual
  models that are advantageous given the context (language, syntax,
  surrounding code). Certainly prior research demonstrates novices and experts both have resources for creating, evaluating, and adapting their own conceptual metaphors to suit the context of the problem \cite{hammer_static_2011,izsak_we_2003,izsak_students_2004}
\item
  To model how programmers think with conceptual models, a suitable
  framework should be able to account for the dynamic, context-sensitive
  deployment of conceptual knowledge.
\item
  To model how programmers develop expertise, a suitable framework
  should be able to describe higher-order phenomena. Such phenomena
  include explaining how programmers come to have conceptual models or
  generate new ones, why they decide to deploy them, and how programmers
  consider which conceptual model (i.e., which way to think about code)
  is appropriate.
\end{enumerate}

@@

\subsection{Manifold models of cognition explain context-dependence
and the growth of
expertise}\label{manifold-models-of-cognition-explain-context-dependence-and-the-growth-of-expertise}

@@

The authors show that we might think of a rubber band as a different
conceptual entity depending on context. In several different
situations---wrapped around a newspaper, pulled taut as a string, spun
to store energy in a toy plane propeller---we intuitively think about
the rubber band's physical behavior differently: one as a negligible
part of the newspaper's point mass, another as a transverse pendulum
(likely) obeying Hooke's Law, and the third as a torsional spring. Those
differences in intuitive thinking reflect the contextual dependency of
what we know about the physical world:

@@

An example helps ground this in-pieces approach to epistemology. Russ, Coffey, Hammer, and Hutchison, describe the
situation where an elementary student reasons about why an empty juice
box collapses when you suck on the straw \cite{russ_making_2008}. One student gives what the
authors deem to be an excellent mechanistic account of why the juice box
collapses. But, as the teacher seems to steer the discussion toward
vocabulary---in this case, ``pressure''---the student clearly pulls back
from her mechanistic reasoning, seems much more diffident, and claims
that pressure is hard to explain. That example highlights the disconnect
between doing science as knowing vocabulary and doing science as
reasoning mechanistically. Moreover, it strikingly highlights that a
student who by all accounts produced an excellent explanation of how
pressure works was left nonetheless with the impression that pressure
was hard to explain.\footnote{In fairness to the teacher, I'm trying to
  focus on the result of the interaction and far less on the intent the
  teacher might have had. The student still left with a sense that
  science might be about vocabulary, even if that's not the view the
  teacher would espouse or was trying to enact in the moment.}

>>>>

\section{Abstract}\label{Abstract}

@@

In 1996, learning scientist David Hammer argued ``Not all thoughts students express need to be understood as directly reflecting stable, stored knowledge structures. What the misconceptions perspective treats as a stored construct may alternatively be treated as an act of construction.'' \cite{hammer_misconceptions_1996}. In the 20 years since, researchers---particularly in science education---have pushed cognitive models beyond the static assumptions of classical misconception accounts of knowledge. But, a striking preponderance of computing education research models cognition via misconceptions. Such research assumes students have entrenched, wrong ideas about computation. And, while misconceptions research in computing education has advanced our understanding of teaching and learning, its narrow also begets problems growing too big to ignore:

\begin{enumerate}
  \tightlist
  \item When we use misconceptions to treat student ideas as right or wrong, we establish and perpetuate a deficit model of student understanding.
  \item The more classical misconceptions-basedresearch we publish as a community, the stronger that deficit narrative becomes in our discourse.
\end{enumerate}

In this paper, we aim to start remedying those problems. We first reanalyze historical data using a knowledge-in-pieces perspective to show how we might recover accounts of students' productive knowledge. We then argue through example that expertise in programming can involve the fluid and active construction of conceptual metaphors to deal with idiosyncrasies within and across programming languages. These examples help broaden our sense of applicable cognitive theory by showing how distributed cognition can be a powerful framework for analyzing activity. Finally, we conclude with an exploration of epistemological concerns, arguing that a fundamental concern for our research community should be paying attention to what counts as knowledge and knowing in computing learning environments.

@@

To be clear, it's not just that graphics might improve our efficiency in
debugging a statement like line 11. Rather, applying an interpretive frame that
treats an assignment statement as a kinematic position update lets us
\emph{use conceptual knowledge from physics} to diagnose and fix
problems in code. For a comparative perspective, consider Hutchins's argument that for aircraft pilots, ringing an aircraft speedometer with physical markers transforms the cognition involved in landing a plane:

@@

In Hutchins's \cite{hutchins_how_1995} example, pilots distribute cognition across time (1) by making approach speed calculations mid-flight, well before they're needed. Pilots then distribute those calculations across space (2) by ringing the speedometer with bugs to represent distinct speeds. And, it is this act of distribution that lets pilots use spatial-adjacency perception (visually making sure the needle is on the bugs) during approach instead of having to calculate, compute, and compare speed numbers during approach.

In our kinematics debugging example, cognition gets distributed across time and space (1) by creating a visual trace of the animated object's position (a persistent trail of dots). And, it is this act of dot-tracing that lets a programmer invokve spatial-adjacency perception and conceptual kinematics knowledge (how positions uniformly sampled in time can reflect constant velocity vs. acceleration) that wouldn't seem applicable if the reassignment statement were stripped of its context.

the cognitive act of debugging in this instance.    There,  For comparison, consider

@@

\subsubsection{Example 2: The Plus Sign Means Lots of Things}

@@

The catch here is that the plus operator takes on different roles in
different languages because of how those languages define its use.
Statement 1 increments a number in C; Statement 2 appends the letters
``ly'' to a string; Statement 3 composes a layer of points onto a statistical
graphics plot. These different kinds of operations become even more
apparent and consequential when, for example, such statements are
repeated in the same language within the same file, as we'll see in Example 3.

\subsubsection{Example 3: The Same Symbol Template \emph{Within} The Same Language Can Still Mean Different Things}

In the R/ggplot2 code below, the author uses multiple
reassignment statements to compose a statistical graphics plot.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Programming can be helped by applying conceptual models to code,
  particularly when relevant domain-knowledge structures can
  advantageously transform a problem (example 1)
\item
  But, conceptual models don't work all the time for all statements.
  Because languages are designed differently, the same syntax can
  actually correspond to very different operations in code (example 2).
  And, that's true both within and across languages (example 3).
\item
  Consequently, it makes less sense to treat conceptual models as right
  or wrong, and more sense to treat them as differentially advantageous
  for thinking about what a piece of code does. (Thinking a ``+''
  implies numerical addition isn't \emph{globally wrong} in JavaScript,
  but it won't explain why 1 + ``1'' yields ``11'' as a result.)
\item
  It seems plausible that successful programmers, when reasoning about
  or writing code, are able to dynamically access or deploy conceptual
  models that are advantageous given the context (language, syntax,
  surrounding code). Certainly prior research demonstrates novices and experts both have resources for creating, evaluating, and adapting their own conceptual metaphors to suit the context of the problem \cite{hammer_static_2011,izsak_we_2003,izsak_students_2004}
\item
  To model how programmers think with conceptual models, a suitable
  framework should be able to account for the dynamic, context-sensitive
  deployment of conceptual knowledge.
\item
  To model how programmers develop expertise, a suitable framework
  should be able to describe higher-order phenomena. Such phenomena
  include explaining how programmers come to have conceptual models or
  generate new ones, why they decide to deploy them, and how programmers
  consider which conceptual model (i.e., which way to think about code)
  is appropriate.
\end{enumerate}

@@

\subsection{Manifold models of cognition can explain context-dependence
and the growth of
expertise}\label{manifold-models-of-cognition-explain-context-dependence-and-the-growth-of-expertise}

@@

The authors show that we might think of a rubber band as a different
conceptual entity depending on context. In several different
situations---wrapped around a newspaper, suspending a bob weight, pulled taut as a string, spun
to store energy in a toy plane propeller---we intuitively think about
the rubber band's physical behavior differently: one as a negligible
part of the newspaper's point mass, two different kinds of harmonic pendulums, and finally a torsional spring. Those
differences in intuitive thinking reflect the contextual dependency of
what we know about the physical world:

@@

An example helps ground this in-pieces approach to epistemology. Russ, Coffey, Hammer, and Hutchison, describe the
situation where an elementary student reasons about why an empty juice
box collapses when you suck on the straw \cite{russ_making_2008}. One student gives what the
authors deem to be an excellent mechanistic account of why the juice box
collapses:

\begin{quote}
  In explaining the phenomenon, Erin focuses on the role played by the air located inside the juice box. She describes the air inside as actively pushing out on all sides of the box holding them out and flat. When that air is removed from the box (by sucking it out through the straw), there is no longer anything pushing from the inside to hold the box out, so the sides cave in. We call this description of the phenomenon the “inside-pusher” model because of its focus on what happens inside the box. \cite{russ_making_2008}
\end{quote}

But, as the teacher seems to steer the discussion toward canonically correct
vocabulary---in this case, ``pressure''---and Erin clearly pulls back
from her mechanistic reasoning, seems much more diffident, and claims
that pressure is hard to explain. That example highlights the disconnect
between doing science as knowing vocabulary and doing science as
reasoning mechanistically. Moreover, it strikingly highlights that a
student who by all accounts produced an excellent explanation of how
pressure works was left nonetheless with the impression that pressure
was hard to explain. In other words, in that moment what counted as knowing was using the word pressure. Erin's mechanistic explanation---beautiful and complex as it was---didn't use that magic word, and in a possible effort to push the class toward canonical behavior, the teacher shut down Erin's reasoning.
